Dutch, PPs, ʀ-pronouns

### 2019-12-21 paper.tex
zaterdagochtend
werkt niet meer (op overleaf wel zo ongeveer!)
intro is bijna klaar, eerste sectie is halverwege

Mismatching free relatives give a unique insight into the internal structure of r-pronouns.

The puzzle is the following. In mismatching free relatives we sometimes see \tit{met wat} and sometimes \tit{waarmee}.

\ex.
\ag. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'
\bg. *Ik heb gekocht met wat jij schildert.\\
 I have bought with what you paint\\
 `I bought what you are painting with.'

\ex.
\ag. *Ik schilder waarmee jij hebt gekocht.\\
 I paint {where with} you have bought\\
 `I paint with what you bought.'
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'

A few things should be noticed here. First, \tit{met wat} and \tit{waarmee} appear in a complementary distribution. Second, \tit{waar} in \tit{waarmee} means \tit{where}, but \tit{waarmee} does not express a location. Third, the preposition \tit{met} changes into \tit{mee}.

This paper: \tit{waarmee} is the default realization of an instrumental when all features form a constituent. If not, then \tit{met wat}. The \tit{r} in ʀ-pronouns does not necessarily express \tsc{location}, in \tit{waarmee} is only expresses \tsc{thing}. The change of \tit{met} into \tit{mee} is a reflex of stranded features/a different spellout of features.

Unmarked examples are constructed and have been verified by native speakers.

\section{\tit{\tsc{r}}-pronouns as default \\
Distribution between \tit{waarmee} and \tit{met wat}}

Repeating the first example from the introduction. What we see here is embedded clause verb requires its object to be an instrumental and the main clause verb wants an accusative object. \tit{waarmee} is correct weer, \tit{met wat} is not.

\ex.
\ag. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'
\bg. *Ik heb gekocht met wat jij schildert.\\
 I have bought with what you paint\\
 `I bought what you are painting with.'

 \tit{Waarmee} is an ʀ-pronoun, a term introduced by \cite{riemsdijk1978}. ʀ-pronouns are not specific to the inanimate relative pronouns. No inanimate pronominal elements in Dutch is able to act as the object of a preposition. They are all replaced by an ʀ-pronoun. As suggested by the name, ʀ-pronouns always contain the morpheme \tit{r}.

 Dutch has the personal pronouns \tit{haar} `her', \tit{hem} `him' and \tit{het} `it' that can be used as animate and inanimate objects of verbs, as illustrated in \ref{ex:objverb}.

@@

The example in \ref{ex:prepani} shows that for animate objects the same pronouns (\tit{haar} `her' and \tit{hem} `him') appear as objects of prepositions. However, the inanimate personal pronoun \tit{het} `it' cannot be used as an object of a preposition, shown in \ref{ex:prephet}. Instead, an ʀ-pronoun appears. This is illustrated in \ref{ex:preper}.

\ex. Objects of prepositions \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met het.\\
 I paint with a brush\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder er-mee.\\
 I paint \tit{er}-with\\
 `I am painting with it.'\label{ex:preper}

\tit{met} is not the only preposition with which this happens.

@@

@@

What this shows is that ʀ-pronouns are not exceptions, but they are the default. There is something special about the examples in which \tit{met wat} is used. I suggest that the crucial difference lies in constituency. \tit{Waarmee} can only be used if the instrumental PP forms a proper constituent, i.e. a constituent to the exclusion of any other elements. In the examples from xx to xx above this is the case.

@@

>>>>

hoi

The term \tsc{r}-pronoun \citep{riemsdijk1978} refers to a set of nominal elements that can strand prepositions in Dutch (and German). It is special because Dutch is a normally non-preposition stranding language. \tsc{R}-pronouns morphologically consist of a combination of a preposition and locative pronoun. In the example \ref{ex:klimerop} \tit{mee} `with' is the preposition and \tit{er} `there' is the locative pronoun.

\exg. Ik klim er-op.\\
 I climb there-on\\
 `I am climbing on it.'\label{ex:klimerop}

This paper focuses on a single \tsc{r}-pronoun in a specific type of free relative construction, namely on the relative pronoun \tit{waarmee} `with what' in a free relative construction in which the predicates require different cases. The \tsc{r}-pronoun \tit{waarmee} `with what' is interesting for two reasons. First, just like for all \tsc{r}-pronouns, the wh-element is the locative, but there is no meaning component related to location in `with what'. Second, the preposition \tit{met} `with' changes into \tit{mee} `with' when it is combined withn \tsc{r}-pronoun.

The construction I focus on is a free relative construction, in which the two predicates combine with different cases. I illustrate this in \ref{ex:gekochtwaarmee}. The predicate in the embedded clause, \tit{schildert} `paint', combines with an instrumental PP. The predicate in the main clause clause, \tit{gekocht} `bought' combines with an accusative DP. The \tsc{r}-pronoun \tit{waarmee} `with what' is used here.

%%%%%%
Footnote
In this example \tit{mee} `with' is stranded and \tit{waar} `where' is moved to the left edge of the embedded clause. It is also possible for \tit{waar} `where' to bring \tit{mee} `with' along, as in \ref{ex:meealong} but it is regarded as less natural.

\exg. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'\label{ex:meealong}
Footnote
%%%%%%

\exg. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'\label{ex:gekochtwaarmee}

If one were to switch around the predicates between the clauses, the \tsc{r}-pronoun does not appear anymore. In \ref{ex:schildermet}, \tit{schilder} `paint' combines with an instrumental PP in the main clause and \tit{gekocht} `bought' combines with an accusative DP in the embedded clause. The use of an \tsc{r}-pronoun is ungrammatical, as indicated by the ungrammaticality of \ref{ex:schilderwaarmee}. Instead, a combination of the regular instrumental preposition \tit{met} `with' and the regular wh-pronoun \tit{was} `what' in used.

\ex.\label{ex:schildermet}
\ag. *Ik schilder waarmee jij hebt gekocht.\\
 I paint {where with} you have bought\\
 `I paint with what you bought.'\label{ex:schilderwaarmee}
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'\label{ex:schildermetwat}

The use of \tit{met wat} `with what' is ungrammatical in the context in which \tit{waarmee} `with what' appeared in \ref{ex:gekochtwaarmee}. This is illustrated in \ref{ex:gekochtmetwat}.

\exg. *Ik heb gekocht met wat jij schildert.\\
 I have bought with what you paint\\
 `I bought what you are painting with.'\label{ex:gekochtmetwat}

In this paper I show that distribution of \tit{waarmee} `with what' and \tit{met wat} `with what' in these free relative constructions gives us a unique insight into the internal structure of \tsc{r}-pronouns. In what follows I show that \tsc{r}-pronouns and regular preposition compete to spell out the same syntactic features. If all features form a constituent, the \tsc{r}-pronoun surfaces. If the constituent is interrupted, the preposition-pronoun combination shows up. This straightforwardly follows in a system in which spellout targets phrasal constituents: Nanosyntax \citep{starke2006}.

This paper is structured as follows. First, I show that it really is constituency. Then I decompose \tsc{r}-pronouns, prepositions and regular pronouns. Last, I show in derivations that constituency connects to the choice for \tsc{r}-pronoun or preposition and regular pronoun. Unmarked examples are constructed and have been verified by native speakers.

\section{\tsc{R}-pronouns as default}

In the introduction I discussed the distribution between \tit{waarmee} `with what' and \tit{met wat} `with what' in free relatives with predicates that combine with different cases. Table \ref{tbl:distribution} repeats the generalization. When the main clause predicate combines with an accusative and the embedded clause predicate with an instrumental, \tit{waarmee} is grammatical and \tit{met wat} is ungrammatical. When the main clause predicate combines with an instrumental and the embedded clause predicate with an accusative, \tit{waarmee} is ungrammatical and \tit{met wat} is used.

\begin{table}[ht]
	\center
	\caption {Distribution between \tit{waarmee} and \tit{met wat}}
	\begin{minipage}{0.27\linewidth}
		\begin{tabularx}{\textwidth}{ccc}
		\toprule
                              & \tit{waarmee} & \tit{met wat} \\
		\midrule
    m:\tsc{acc}, e:\tsc{ins}  & ✓             & *             \\
    m:\tsc{ins}, e:\tsc{acc}  & *             & ✓             \\
    \bottomrule
\end{tabularx}
\end{minipage}
\end{table}\label{tbl:distribution}

The goal of this section is to show that \tit{waarmee} is the default as instrumental relative pronoun. In order to do that, I discuss the distribution of \tsc{r}-pronouns and regular pronouns in more general. Dutch has the personal pronouns \tit{haar} `her', \tit{hem} `him' and \tit{het} `it' that can be used as animate and inanimate objects of verbs, as illustrated in \ref{ex:objverb}.

@@

The example in \ref{ex:prepani} shows that for animate objects the same pronouns (\tit{haar} `her' and \tit{hem} `him') appear as objects of prepositions. However, the inanimate personal pronoun \tit{het} `it' cannot be used as an object of a preposition, shown in \ref{ex:prephet}. Instead, an ʀ-pronoun appears. This is illustrated in \ref{ex:preper}. \ref{erprep} shows that the \tsc{r}-pronoun obligatorily moves to the left of the pronoun.

\ex. Objects of prepositions \label{ex:objprep}
\ag. Ik schilder samen met haar/hem.\\
 I paint together with her/him\\
 `I am painting together with her/him.'\label{ex:prepani}
\bg. *Ik schilder met het.\\
 I paint with a brush\\
 `I am painting with it.'\label{ex:prephet}
\bg. Ik schilder er-mee.\\
 I paint there-with\\
 `I am painting with it.'\label{ex:preper}
\bg. *Ik schilder mee er.\\
 I paint with-there\\
 `I am painting with it.'\label{ex:erprep}

\tit{Met} is not the only preposition with which this happens. \tit{Op} `on' and \tit{in} `in' do not combine with the inanimate personal pronoun \tit{'t}, but the \tsc{r}-pronoun is used obligatorily.

@@

A picture similar to the one for inanimate personal pronouns appears with wh-inanimates.

object of verbs \tit{wat}.
object of preposition with \tit{wat}

then show they appear in the contexts below.

@@

In sum, no inanimate pronominal elements in Dutch is able to act as the object of a preposition. It is always the \tsc{r}-pronoun that appears. What this shows is that ʀ-pronouns are not exceptions, but they are the default. There is something special about the examples in which \tit{met wat} is used. I suggest that the crucial difference lies in constituency. \tit{Waarmee} can only be used if the instrumental PP forms a proper constituent, i.e. a constituent to the exclusion of any other elements. --explain here how they all form a constituent. In the examples from xx to xx above this is the case.

@@

\ex.
\ag. Ik heb gekocht waar jij mee schildert.\\
 I have bought where you with paint\\
 `I bought what you are painting with.'
\bg. Ik schilder met wat jij hebt gekocht.\\
 I paint with what you have bought\\
 `I paint with what you bought.'
